Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-26 Manuscript/TEM_draft.tex
Edits to abstract and title page

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$, and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland;\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland;}\\
\end{center}
\medskip
\noindent{\bf Corresponding author:} Thomas Guillerme, School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.\\
\vspace{1in}

@@

\newpage
\begin{abstract}
Living species represent only a small proportion of all species that have ever lived. Ignoring fossil taxa may therefore lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology. This fact has led to an increasing consensus among scientists that both living and fossil taxa must be included in macroevolutionary studies. One approach called Total Evidence, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both living and fossil taxa at the tips. Although the Total Evidence approach seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

In this study we assess the effect of missing data on tree topologies inferred from Total Evidence matrices. Using simulations we investigate three major factors that directly affect the completeness of the morphological part of the matrix: (1) the proportion of living taxa with no morphological data, (2) the amount of missing data in the fossil record, and (3) the overall number of morphological characters in the matrix.

% NC: This needs changing to reflect what we decided to focus on, i.e. our recommendations. We don't want to lead with the methods thing. The two previous paragraphs are fine and just need tightening up a bit
%We find that, when using a clade conservative metric such as Robinson-Foulds distance, Bayesian consensus trees recovers the right topology better than Maximum Likelihood or than a subset of Bayesian posterior tree distributions and this regardless of the amount of missing data (minimal tree similarity of 0.69 in the worst scenario). Regarding the data, good topological recovery is not related to the amount of missing data \textit{per se} but to the amount of data overlap. The two factors that influences the most this data overlap are the parameters 1 (proportion of coded living taxa) and 3 (number of morphological characters).

We find that increasing the number of living taxa with morphological characters (1) and the overall amount of morphological data (3) have a greater impact on the ability to recover the "correct" topology than increasing the data completeness in the fossil record (2). Therefore, we suggest that sampling effort should be focusing on these two first aspects to increase the accuracy of topology in the Total Evidence framework. Additionally, we found that the Bayesian consensus tree consistently recovers topology that are closest to the "correct" topology than any other method. Therefore, we also suggest, after the data collection effort, that one should fix the tree topology using the Bayesian consensus tree topology for any further tree inference or macroevolutionary analysis.

>>>>

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$$^*$, and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.}\\
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{Zoology Building, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.}\\
\vspace{1in}

@@

\newpage
\begin{abstract}
To fully understand macroevolutionary patterns and processes, we need to include both extant and extinct species in our models. This requires phylogenetic trees with both living and fossil taxa at the tips. One way to infer such phylogenies is the Total Evidence approach which uses molecular data from living taxa and morphological data from living and fossil taxa. Although the Total Evidence approach is very promising, it requires a lot of data and is therefore likely to suffer from missing data issues that may affect its ability to infer correct phylogenies.

Here we use simulations to assess the effects of missing data on tree topologies inferred from Total Evidence matrices. We investigate three major factors that directly affect the completeness of the morphological part of the matrix: the proportion of living taxa with no morphological data, the amount of missing data in the fossil record, and the overall number of morphological characters in the matrix. We infer phylogenies from complete matrices and from matrices with various amounts of missing data, and then compare missing data topologies to the "best" tree topology inferred using the complete matrix.

We find that the number of living taxa with morphological characters and the number of morphological characters in the matrix, are more important than the amount of missing data in the fossil record for recovering the "best" tree topology. Therefore, we suggest that sampling effort should be focused on morphological data collection for living species to increase the accuracy of topological inference in a Total Evidence framework. Additionally, we find that Bayesian methods consistently outperform other tree inference methods. We therefore recommend using Bayesian consensus trees to fix the tree topology prior to further analyses such as Tip-Dating.
